---
title: "Wolf observations, rough intensity vol. 2.8 (overdispersion, m2 area, wolf year)"
author: "Tuomas Rajala"
date : "`r Sys.Date()`"
output:
  html_document:
    toc: true
    code_folding: hide
    fig_caption: true
    number_sections: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(sf)
library(ggplot2)
library(dplyr)
library(tidyr)
library(raster)
library(mgcv)
library(parallel)
theme_set(theme_bw())
knitr::opts_chunk$set(fig.width=9, fig.height=3, echo=TRUE, message=FALSE, warning=FALSE)
```

* v2.8: look at overdispersion; time in wolf years; homogenise spatial units; check if "main" effect `ti` terms make difference

* v2.7: skipped to avoid numbering issues from before VC

* v2.61: github version; should be the same model result

* v2.6: Include data processing and aggregation steps, excl. raw-tassu compilation from excels and gps-territory geometry calculations.

* v2.5: Drop data before 2020 to allow prediction quality assessment

* v2.4: Drop the `ryhmassa == TRUE` condition, data more model roughly doubled.

* v2.3: Include Jan and Feb with March in the active territories

* v2.2: Offset so that $E y_i = \lambda_i / A_{(i)}$, i.e. as intended it is harder to observe packs with larger territories.

* v2.1: Only those obs for which `(ryhmassa == TRUE | havainto_id_count > 1)  &  (havainto_id_subidx == 1)`, meaning one row of data is a unique observation of a pack, no matter the size as long as there are more than 1 wolves per registered observation. 

* v2.0: one row of data is a registered observation in the tassu. Could be repeated for many wolves, could be a of a single wolf. Note that big packs have many datapoints per observation coordinate. Also, we add the area of the territory each counting cell hits as offset.

# Intro

Model the intensity of wolf-observations in  the Tassu data, conditioned on "wolf present" subset of Tassu that hits active wolf territories. Territories derived as the convex hulls of GPS-collar tracking data and the expert determined territories on Mays of 2017-2019. 


# Data

```{r}
####################################################
# Region of interest:
#
crs1 <- 3067 # use this for all
crs1_c <- paste0("EPSG:", crs1)
finland <- readRDS("../data/finland.rds")
finland <- st_transform(finland, crs1)
# Drop North from geometries. Hard cut, no other preference given.
box1 <- st_bbox(c(xmin=0, xmax=100000000, ymin=0, ymax = max(72e5 + 30000)), crs=st_crs(finland))
finland1 <- st_crop(finland, box1)

# 
agg_out_file <- "processed/celdat-v2.6.rds"

celdat1 <- readRDS(agg_out_file)
# for plots
dat2 <- readRDS("processed/tassu_dat2-v1.rds")
terr <- readRDS("processed/territories-v1.rds")

# for prediction
corine8s <- readRDS("processed/corine8s-v1.rds")
droad0 <- crop(stack("../data/digiroad_luokka12_1x1km_2021-04-14.grd"), 
               as_Spatial(finland1))
droad1 <- setNames(droad0, "droad.cl12")
```

Dataframes prepared in `1-aggregate-model-data-frame-v1.R`.


## Some plots

Tassu observations 

```{r}
addmargins( table(dat2$laji, dat2$vuosi) )
```

Tassu wolf on map, spatial

```{r, fig.height=7, fig.width=9}
p <- dat2 %>% 
  ggplot() + 
  geom_sf(data = finland1, fill=NA) + 
  geom_sf( data = terr, aes(fill = type), alpha=.3, col = NA)+
  geom_point(aes(x,y), alpha=0.2, size = 0.2) + 
  ggtitle("Tassu wolf | GPS+expert territories ")  

p
```



```{r}
dat2 %>% ggplot() + geom_histogram(aes(timestamp_utc), bins = 300)
```


Number of cells: 

```{r}
celdat1 %>% ggplot() + geom_bar(aes(kkf)) + facet_wrap(~vuosi, scale="free_y") + 
  ggtitle("Number of active cells")
```

```{r}
print(c(cells_total = nrow(celdat1), cells_with_wolves = sum(celdat1$nsusi > 0)))
```



Note changing scale.

Number of wolf reports in those cells:

```{r}
celdat1 %>% group_by(vuosi, kkf) %>% summarise(nsusi =sum(nsusi))  %>% 
  ggplot() + geom_col(aes(kkf, nsusi)) + facet_wrap(~vuosi, scale="free_y") + ggtitle("Number of wolf reports")
```

# Modeling details

Model is Poisson counting, in cells of 1 km x 1km x 1 month.

The model is for the count of wolf observations in a cell $i$ of size $V_i$

$$y_i |\lambda_i \sim Poisson(\lambda_i / A_{(i)})$$
$$\log \lambda_i = \tilde{year}_i + \tilde{month_i} + s(x_i, y_i) + corine\{2,...,8\}_i + forestroad_i$$
where $\sim$ means factors and

* $A_{(i)}$ is the area of the active territory observation cell $i$ hits  (offset, idea: Larger area -> harder to spot the pack; note we use the sum of all areas that might overlap there, so not strictly the union)
* $s(x,y)$ is a spatially smooth term (`gam::ti`, smoothness via GCV)
* $corinej_i$ is $j$th corine-covariate at cell $i$, taken from the 1km x 1km aggregated corine with extra 3km smoother applied (Gaussian blur), and then normalised so each corine cell holds a probability vector. Due to this sum-to-1 linear dependency, the first class is dropped.
* $forestroad_i$ is an amount of forest road 

Note: Data before 2020 used for fitting.



```{r fit-poisson-bam, cache=TRUE}
# final preparations and filterings.
model_dat <- celdat1 %>% 
  filter(vuosi < 2020)

###
# The define model


cn <- colnames(model_dat)
cona <- cn[grepl("^c18", cn)][-1] ## drop first due since simplex, same predictive power as log-ratios
#cona <- cn[grepl("^lr.c18", cn)] ## log-ratios?
cona <- c(cona, "droad.cl12")
corv <- as.formula(paste( ".~ . + ", paste(cona, collapse = " + ")) )

#p0 <- nsusi ~ -1  +  vuosif + kkf  + special_may + ti(x, y) 
p1 <- nsusi ~ -1  +  wolf_yearf + wolf_monthf  + special_may + 
  te(x,y) + #ti(x) + ti(y) + ti(x, y) + 
  offset(-log(area_m2))
#p1s <- nsusi ~  vuosif + kkf  + special_may + offset(-log(area_km2))

#pml0 <- update.formula(p0,  corv)
pml1 <- update.formula(p1,  corv)
#pml1s <- update.formula(p1s,  corv)
# fit
cl <- beginCluster(detectCores()/2)

#t0 <- system.time( pit0 <- bam( pml0, data = model_dat, family = "poisson", cluster=cl) )
t1 <- system.time( pit1 <- bam( pml1, data = model_dat, family = "poisson", cluster=cl) )
#t1s <- system.time( pit1s <- bam( pml1s, data = model_dat, family = "poisson", cluster=cl) )




#print(anova(pit0, pit1))
```



```{r}
phihat <- sum(residuals( pit1, type = "pearson" )^2) / df.residual(pit1)
```

The overdispersion estimate is $\hat\phi \approx `r round(phihat, 2)`$.

Estimated model:

```{r}
sum1 <-  summary(pit1, dispersion = phihat)
sum1
```

Time unit is 1 "month" = 1/12th of a year. 

The baseline is "month = 1", for which the estimate is $log(\lambda)=0$. All other info adds to that. 

# The components of lambda in space and time

Yearly effects:


```{r}
coefs <- sum1$p.table
cnames <- rownames(coefs)
vix <- grepl("year", cnames)
S <-  sum1$cov.scaled
vixS <- grepl("yearf", rownames(S))
vS <-  S[vixS, vixS]
AM <- t(rep(1, sum(vix)))/sum(vix)

out <- list(year_effect = coefs %>% as_tibble(rownames="coef") %>% filter(grepl("yearf", coef)))
out$year_effect %>% mutate(coef = gsub("wolf_yearf",  "", coef), estimate = Estimate, sd = `Std. Error` ) %>%
  ggplot(aes(coef, estimate))+ 
  geom_pointrange(aes(ymin = estimate - 2*sd, 
                      ymax = estimate+2*sd)) + 
  theme(axis.text.x = element_text(angle=30, hjust=1))

```

Compute the average year effect:

```{r}
mean_year_effect <- c(estimate =  AM%*%coefs[vix,1], sd = sqrt( AM%*%vS%*%t(AM) ) )  
print(mean_year_effect)

out$mean_year_effect <- mean_year_effect
```

Then monthly effects follow. Note month=1 was the baseline = 0. 

```{r}
kix <- grepl("monthf", cnames)
out$month_effects <- coefs[kix,1:2] %>%  as_tibble() %>% setNames(c("estimate", "sd")) %>%
  mutate(wolf_month = sprintf("%02i", 2:12))
print(out$month_effects)

out$month_effects %>%  
  ggplot(aes(wolf_month, estimate))+ 
  geom_pointrange(aes(ymin = estimate - 2*sd, 
                      ymax = estimate+2*sd))

```

Spatial: We have two separate spatial terms since Corine-info is also spatial. Compute in south finland.

The predictions include offset = -log[ median($A_{(i)}$) ] i.e. resulting values are computed conditionally that each pixel is included in an average sized territory. 

```{r}
# over the target area, make a dummy dataframe
them <- !is.na(corine8s$Tiet[])
c18df <- data.frame(c18=as.data.frame(corine8s), 
                    as.data.frame(droad1),
                    wolf_yearf = "2010-2011", 
                    wolf_monthf = "1", 
                    special_may = FALSE, # need for predict-method
                    area_m2 = median(model_dat$area_m2),
                    areasum = median(model_dat$areasum)
                    ) %>% 
  bind_cols( coordinates(corine8s) %>% data.frame() )
c18df <- c18df[them,]

cat(sprintf("Median territory size used in prediction: %f km^2\n", c18df$areasum[1]))

fitbest <- pit1
conabest <- cona

# checks
# pred0 <- predict.bam(fitbest, c18df, 
#                       type = "link", 
#                       se.fit = TRUE, 
#                       cluster = cl) 
# pred1  <- predict.bam(fitbest, c18df, 
#                       type = "terms", 
#                       se.fit = TRUE, 
#                       cluster = cl)
# # check
# all.equal(c(pred0$fit), rowSums(pred1$fit), check.attributes=F)
#

pred_corine <- predict.bam(fitbest, c18df, terms = cona, 
                      type = "terms", 
                      se.fit = TRUE, 
                      cluster = cl) 

pred_s <- predict.bam(fitbest, c18df, 
                      terms = "te(x,y)",
                      #terms = c("ti(x)", "ti(y)", "ti(x,y)"), 
                      type = "terms", 
                      se.fit = TRUE, 
                      cluster = cl) 

lammap_smooth <- data.frame(
  log_lam = rowSums(pred_s$fit),
  log_lam.se = sqrt( rowSums(pred_s$se.fit^2)), 
  term = "smooth(x,y)") %>%
  bind_cols(c18df %>% dplyr::select(x, y) ) 

lammap_corine <- data.frame(
  log_lam = rowSums(pred_corine$fit),
  log_lam.se = sqrt( rowSums(pred_corine$se.fit^2) ), 
  term = "corine&digiroad") %>%
  bind_cols(c18df %>% dplyr::select(x, y) ) 

lammaps <- rbind(lammap_smooth, lammap_corine)

# and together
lammap <- data.frame(
  log_lam = rowSums(pred_corine$fit) + rowSums(pred_s$fit),
  log_lam.se = sqrt( rowSums(pred_corine$se.fit^2 ) + 
                     rowSums(pred_s$se.fit^2 ) ), 
  term = "smooth(x,y) + corine&digiroad") %>%
  bind_cols(c18df %>% dplyr::select(x, y) ) 


```

check plot

```{r}
mapit <- function(d) d %>%
  pivot_longer(log_lam:log_lam.se) %>%
  ggplot(aes(x, y)) + 
  geom_raster(aes(fill = value)) +
  coord_fixed() + 
  scale_fill_distiller(palette = "Spectral") + 
  facet_wrap(~term + name)

lammaps %>% filter(term == "smooth(x,y)") %>% mapit()
```

```{r}
lammaps %>% filter(term != "smooth(x,y)") %>% mapit()
```


```{r}
lammap %>% mapit()
```

# Store
save in a  list-of-objects

```{r}
out$spatial <- lammaps
out$model_summary <- sum1
saveRDS(out, "processed/poisson_regression_log_lambda_terms-v2.8.rds")

# store now also the model object
saveRDS(pit1, "processed/poisson_regression_bamfit-v2.8.rds")

```


```{r}
endCluster()
```

